# Evaluation of activity data


## Data reading and processing  
To enable a weekday-specific evaluation, a column with corresponding weekdays for every date was added.

```{r, echo = TRUE, warning = FALSE, message = FALSE}
library(ggplot2)
library(data.table)
library(dplyr)
library(tidyr)
library(lubridate)

activity.data <- setDT(read.csv("activity.csv"))
activity.data$weekday <- wday(activity.data$date,label = TRUE,abbr = FALSE)
```

## Check for missing data  
The data was visualized as heat map to check for missing data. Missing data is colored in grey.

```{r, echo = TRUE, fig.height=10, fig.width=10, warning = FALSE}
plot_missing <- 
        ggplot(data = activity.data, aes(x = interval, y = date, fill = steps)) +
        geom_tile() +
        theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), 
              panel.background = element_rect(fill = "white", color = "white")) +
        facet_wrap(vars(weekday))
plot_missing
```

Data for whole days is missing from the data set.Several weekdays are affected by missing data.
The strategy for imputing the missing data is developed based on an analysis of the steps taken per five minute interval.

## Analysis of steps taken per 5 minutes intervals  
The mean of steps taken at each interval was calculated and the result was plotted as time series.

```{r, echo = TRUE, warning = FALSE}
intervals <- activity.data %>%
        group_by(interval) %>%
        summarise(steps_per_interval = mean(steps, na.rm = TRUE))
interval_plot <- 
        ggplot(data = intervals, aes(x = interval, y = steps_per_interval)) +
        geom_line()
interval_plot

max_steps <- max(intervals$steps_per_interval, na.rm = TRUE)
max_interval_dt <- intervals[intervals$steps_per_interval == max_steps, ]
max_interval <- unlist(max_interval_dt[1,1])
```

The average of steps taken per five minute interval fluctuates between 0 and `r max_steps`. The interval with the highest average of steps is: `r max_interval`

### Influence of weekday on steps taken per 5 minutes intervals  
The calculation of mean steps per 5 minutes intervals was recalculated separately per weekday.

```{r, echo = TRUE, warning = FALSE, message = FALSE, fig.height=10, fig.width=10,}

weekday_summary <- activity.data %>%
        group_by(interval, weekday) %>%
        summarise(mean_steps_per_interval = mean(steps, na.rm = TRUE),
                  median_steps_per_interval = median(steps, na.rm = TRUE))
colors <- c("mean" = "blue", "median" = "red")

weekday_plot <-
        ggplot(data = weekday_summary) +
        geom_line(aes(x = interval, y = mean_steps_per_interval,  
                  color = "mean")) +
        geom_line(aes(x = interval, y = median_steps_per_interval, 
                  color = "median")) +
        labs(y = "steps per interval", color = "plotted value") +
        scale_color_manual(values = colors) +
        facet_wrap(vars(weekday))
weekday_plot
```

The time course of mean steps taken per interval is different for each weekday.Furthermore, the median of steps taken per day is very different from the mean value, indicating a skewed distribution.

## Analysis of steps taken per day  
The sum of steps taken each day was calculated and the result was plotted as histogram.
Mean and median number of steps taken each day were calculated.

```{r, echo = TRUE, warning = FALSE}
steps <- activity.data %>%
        group_by(date) %>%
        summarise(sum_steps_per_day = sum(steps))

steps_per_day <- hist(steps$sum_steps_per_day, col = "lightgray", plot = TRUE, 
                      xlab = "sum of steps per day", 
                      main = "Histogram of steps per day")

median_steps_per_day <- quantile(steps$sum_steps_per_day, probs = 0.5, na.rm = TRUE)
mean_steps_per_day <- mean(steps$sum_steps_per_day, na.rm = TRUE)
```

The count of steps taken per day appears to be normally distributed.  
The median of steps taken per day is `r median_steps_per_day`, the mean is `r mean_steps_per_day`.

### Imputing missing values  
Missing values will be replaced by the median per interval for the specific weekday.A new histogram including the imputed data is shown.

```{r, echo = TRUE, warning = FALSE}
weekday_vector <- unique(activity.data$weekday)
imputed_missing <- as.data.table(matrix(nrow =0, ncol = 4))
colnames(imputed_missing) <- names(activity.data)
for (i in weekday_vector) {
        temp_dt <- activity.data[weekday == i,]
        temp_dt[, steps:= ifelse(is.na(steps), mean(steps, na.rm = TRUE),steps),
                      by = interval]
        imputed_missing <- rbind(imputed_missing, temp_dt)
}  
steps2 <- imputed_missing %>%
        group_by(date) %>%
        summarise(steps_per_day = sum(steps))

steps_per_day2 <- hist(steps2$steps_per_day, plot = TRUE,
                       xlab = "sum of steps per day", 
                       main = "Histogram after imputing missing values")

```

 
